Working document to try to explain why/how ATAs influence coexistence in an intuitive way for referee 3

### 2023-07-28 paper/ExplanationOfATAEffectsWorkingDoc.tex
merge

\subsection{Storage effect and how ATAs can change it}
The storage effect stably contributes to coexistence through 
temporal niche partitioning with EC covariance and buffered population growth.
Under invader-resident analysis [rare-common], EC covariance limits the resident by 
coupling high environmental responses with high intraspecific competition. The magnitude
of the storage effects then depends on the EC covariance of the invader. Because the 
invader is introduced into the resident steady state community at close to zero density, 
its C is only a result of interspecific competition from the resident. Thus, the more closely 
related the invader's EC relationship is, the more interspecific competition it experiences, 
destabalizing coexistence. Equivalently, invader EC relationships with more spread stably
contribute to coexistence (through the storage effect). As seen in Fig. \ref{thefig}, ATAs in 
the invader EC relationship changes its bivariate distribution to both have more spread 
about the "central line" and be more densely situated about the "central line".  
%is central line like niche overlap line? species association line?  
When the invader's EC relationship has more spread (more density in values far away from 
the central line) than what the invader's EC relationship would be if it was symmetric, 
the ATA effect is positive. More specifically, points to the right of the central line are where environmental response is high and competition is low. When points fall in this plane, is when species can "store" the beneficial effects of a good environment to counteract when conditions are less ideal. Thus when the EC relationship is less spread, the invader is less able to recover from rare because the points are far away from the beneficial extremes. 
I think this related to the subadditivity criteria of storage effect, i.e., buffered population growth which is related to nonlinear averaging. Subaddivity/ buffered popualtion growth is when species growth rate responses to competition are less affected/sensitive to bad environmental conditions and more affected and sensitive when conditions are good. So it is nonlinear. So points in the beneficial extremes are when the invader can take advantage of the nonlinear averaging that comes with buffered population growth. ATA'd EC relationships have different distribution of points compared to one that is symmetric with the same marginals and correlation so the asymmetry changes whether or not points are in the beneficial extremes.

@@

\subsection{Dan thoughts 2023 07 26}

% %Attempt 1 - not as good, Dan thinks. May be slightly misleading. See below for attempt 2.
% %
% %I have in mind that the following paragraph will go i the Discussion, with maybe a sentence
% %or two in the Intro as a preview of coming attractions.
% Just as for storage effects, temporal niche differentiation is necessary in order for ATA
% effects to materialize. This makes intuitive sense because temporal niche differentiation is 
% also a prerequistite of storage effects [\cite{Chesson_1994} and Box 2 of \cite{StumpVasseur2023}],
% and ATA effects are a part of storage effects, but see
% SI section X for a detailed explanation. 
% %Topic sentence follows. Possibly we can state something like this in the Intro and say
% %we explain it in the Discussion after formalizing the extent to which ATAs are important.
% If temporal niche differentiation occurs, ATA effects essentially capture the influence of asymmetries 
% in the extent to which storage effects can occur in environmentally good versus bad years,
% as we here explain.
% %
% These asymmetries can make a big difference to total storage effects and coexistence,
% as some of our analyses showed.
% %
% If temporal niche differentiation occurs
% but is partial (as will be typical), the quantities $E$ and $C$ experienced by a currently scarce 
% species (i.e., $E_i$ and $C_{i \bs i})$ will be positively but imperfectly correlated. 
% %
% Storage effects can occur if some buffering mechanism reduces losses in bad years so that
% gains in good years can more than make up for them. 
% %
% Bad years happen whenever the negative effects of $C$ outweigh any 
% positive effects of $E$; but this can occur
% both when 1) $E$ is moderately favorable but $C$ is very strong, as well as when 2) $E$ is very unfavorable,
% or moderately unfavorable and $C$ is not too weak.
% Conversely, good years happen whenever positive effects of $E$ outweigh negative effects of $C$;
% but this can occur when 1) $E$ is very favorable, or moderately favorable and $C$ is not too strong,
% as well as when 2) $E$ is moderately unfavorable but $C$ is very weak. 
% %
% ATA effects essentially quantify the asymmetry of storage that occurs when 
% $E$ and $C$ are both higher than average (the cases 1, above); versus storage that occurs 
% when $E$ and $C$ are
% both lower than average (the cases 2, above). 
% %
% The distinction is an important one because it corresponds to temporal niche differentiation
% which occurs under two very different sets of environmental conditions, possibly via
% different ecological or physiological mechanisms.
% %
% Details, applications, and examples of this intuition 
% are in SI section X, and these ideas are also used there to provide 
% intuitive explanations for the relative differences in the importance
% of ATA effects between our log-normal and beta fecundities lottery models.
% %DAN: You ought to also mention explaining differences with the diatom system,
% %if you figure out an explanation.

%Attempt 2 below - better than the above, but may need to be simplified a bit more,
%if possible
%
Just as for storage effects, temporal niche differentiation is necessary for ATA
effects. This makes intuitive sense because temporal niche differentiation is 
also a prerequistite of storage effects [\cite{Chesson_1994} and Box 2 of \cite{StumpVasseur2023}],
and ATA effects are a part of storage effects, but see
SI section X for a detailed explanation. 
%
Assuming that temporal niche differentiation is present, how do ATA effects then happen?
We here offer an intuitive explanation. 
%
If temporal niche differentiation occurs
but is partial, the $E$ and $C$ experienced by a currently scarce 
species (i.e., $E_i$ and $C_{i \bs i})$ will be positively but imperfectly associated, 
and the association may be asymmetric. 
%
Storage effects can occur if some buffering mechanism reduces losses in bad years so that
gains in good years can more than make up for them. 
%
Bad years happen when the negative effects of $C$ outweigh any 
positive effects of $E$; but this can occur
both 1) when $E$ is moderately favorable but $C$ is very strong, as well as 2) when $E$ is very unfavorable,
or moderately unfavorable and $C$ is not too weak.
Conversely, good years happen when positive effects of $E$ outweigh negative effects of $C$;
but this can occur 1) when $E$ is very favorable, or moderately favorable and $C$ is not too strong,
as well as 2) when $E$ is moderately unfavorable but $C$ is very weak. 
%
It turns out that the potential for storage to occur when 
$E$ and $C$ are both higher than average (the cases 1, above); versus the potential  
when $E$ and $C$ are
both lower than average (the cases 2) are not necessarily equivalent.
Depending on the growth rate function, $r(E,C)$, storage potential may differ
across the spectrum from 1 to 2.
%
ATA effects are the extent to which asymmetries of association between $E$ and 
$C$ are aligned with the potential for storage across the dichotomy 1 versus 2.
A comparison is made to a benchmark of symmetric association between $E$ and $C$.
%
ATA effects therefore help quantify different degrees of temporal niche differentiation 
occuring under different sets of environmental conditions (such as cases 1 and 2), possibly via
different ecological or physiological mechanisms.
%
Details, applications, and examples of this intuition 
are in SI section X, and these ideas are also used there to provide 
intuitive explanations for the relative differences in the importance
of ATA effects between our log-normal and beta fecundities lottery models.
%DAN: You ought to also mention explaining differences with the diatom system,
%if you figure out an explanation.

\subsection{SI section X corresponding to the thoughts above}

Just as for storage effects [see \cite{Chesson_1994} and Box 2 of \cite{StumpVasseur2023}], 
temporal niche differentiation is necessary in order for ATA
effects to materialize; we here explain why in detail. ``Lack of temporal niche differentiation'' can be interpreted
to mean that species respond in identical ways to environmental fluctuations. 
In a two-species model such as those we consider, this should mean a monotonically
increasing relationship between $E_i$ and $C_{i \bs i}$. The former is the environment
as perceived by species $i$. The latter is competition that a scarce invader $i$ experiences
in a monoculture of $j$, and that competition comes only from $j$ (because $i$ itself
is scarce), so should correspond to the environmental response of $j$. If 
$E_i$ and $C_{i \bs i}$ are monotonically related, then $(E_i,C_{i \bs i})$
and $(E_i^\nat,C_{i \bs i}^\nat)$ are distributed in the same way, so 
$\E[r_i(E_i,c_{i \bs i})]-\E[r_i(E_i^\nat,c_{i \bs i}^\nat)]=0$.
Therefore, 
\begin{align}
\Delta_i^{[EC]} &= (\olr_{i \bs i}-\olr_{j \bs i})-(\olr_{i \bs i}^\nat - \olr_{j \bs i}^\nat) \\
&= \olr_{i \bs i}-(\olr_{i \bs i}^\nat - \olr_{j \bs i}^\nat) \\
&= \E[r_i(E_i,C_{i \bs i})]-\E[r_i(E_i^\nat,C_{i \bs i}^\nat)]+\E[r_j(E_j^\nat,C_{j \bs i}^\nat)] \label{eq:expri}\\
&= \E[r_j(E_j^\nat,C_{j \bs i}^\nat)].
\end{align}
But we already know $\E[r_j(E_j^\nat,C_{j \bs i}^\nat)]=0$
because there is no niche differentiation between $j$ and itself, so 
$E_i$ and $C_{j \bs i}$ should be monotonically related. So 
$\E[r_j(E_j^\nat,C_{j \bs i}^\nat)] = \E[r_j(E_j,C_{j \bs i})]=0$, giving
$\Delta_i^{[EC]}=0$. So temporal
niche differentiation is necessary for ATA effects on coexistence to 
materialize.

If temporal niche differentiation occurs but is partial, 
then $E_i$ and $C_{i \bs i}$ will be positively but imperfectly correlated with
each other, corresponding to similar but non-identical responses of species to
environmental fluctuation; but how do ATA effects then happen? We here extend the intuition
introduced in the Discussion and add detail.

Storage can occur if population
increases during good years outweigh decreases during bad years. For instance,
if $r_i(E_i,C_{i \bs i})$ is very large when the balance of $E_i$ and 
$C_{i \bs i}$ is favorable, but only moderately small when this balance is
unfavorable, the expected value $\E[r_i(E_i,C_{i \bs i})]$ in (\ref{eq:expri})
will be positive. But the potential for such storage of population growth
may differ between when $E_i$ and $C_{i \bs i}$ are both high compared
to when they are both low. 
For instance, for the beta fecundities lottery model, $E_i/C_{i \bs i}$, a quantity
which features prominently in $\E[r_i(E_i,C_{i \bs i})]=\log(1-\delta+E_i/C_{i \bs i})$,
takes values bounded above by $\frac{2\eta_i \delta}{\eta_j}$ 
and bounded below by $\frac{\eta_i \delta}{2 \eta_j}$ when $E_i$ and $C_{i \bs i}$
are both in the upper halves of their assumed beta distribtions (recall that 
these quantities were assumed to be distributed, respectively, as $\eta_1 \beta$
and $\eta_2 \beta/\delta$ for $\beta$ a beta-distributed random variable with parameters 
$0.5$ and $0.5$). These bounds are separated only by a factor of four. 
On the other hand, when $E_i$ and $C_{i \bs i}$ are both in the lower halves 
of their distributions, $E_i/C_{i \bs i}$ can take any positive value because
then either $E_i$ or $C_{i \bs i}$ can take values arbitrarily close to $0$. 
Because values of $E_i/C_{i \bs i}$ have the potential to be much larger when 
$E_i$ and $C_{i \bs i}$ are both in the lower halves of their distributions
than when these quantities are both in their upper halves, and because
the growth rate, $\log(1-\delta+E_i/C_{i \bs i})$, buffers low values of 
$E_i/C_{i \bs i}$ (the infemum of the growth rate is $\log(1-\delta)$, whereas the supremum is
infinite), the storage effects contributed by the left tails of 
$E_i$ and $C_{i \bs i}$ have the potential to be much greater than those contributed
by the right tails. Essentially, values of $\log(1-\delta+E_i/C_{i \bs i})$
occurring when $E_i$ and $C_{i \bs i}$ are both in the upper halves of their
distributions can never exceed $\log(1-\delta+\frac{2\eta_i \delta}{\eta_j})$,
whereas values ocurring when $E_i$ and $C_{i \bs i}$ are both in the lower
halves of their distributions can be arbitrarily large (though with decreasing 
probability for larger values).

\bibliographystyle{ecology_letters2}
\bibliography{refs}

>>>>

\subsection{Storage effect and how ATAs can change it}
The storage effect stably contributes to coexistence through 
temporal niche partitioning with EC covariance and buffered population growth.
Under invader-resident analysis [rare-common], EC covariance limits the resident by 
coupling high environmental responses with high intraspecific competition. The magnitude
of the storage effects then depends on the EC covariance of the invader. Because the 
invader is introduced into the resident steady state community at close to zero density, 
its C is only a result of interspecific competition from the resident. Thus, the more closely 
related the invader's EC relationship is, the more interspecific competition it experiences, 
destabalizing coexistence (maybe only true with small variance). Equivalently, invader EC relationships with more spread stably
contribute to coexistence (through the storage effect). As seen in Fig. \ref{thefig}, ATAs in 
the invader EC relationship changes its bivariate distribution to both have more spread 
about the "central line" and be more densely situated about the "central line".  
%is central line like niche overlap line? species association line?  
When the invader's EC relationship has more spread (more density in values far away from 
the central line) than what the invader's EC relationship would be if it was symmetric, 
the ATA effect is positive. More specifically, points to the right of the central line are where environmental response is high and competition is low. When points fall in this plane, is when species can "store" the beneficial effects of a good environment to counteract when conditions are less ideal. Thus when the EC relationship is less spread, the invader is less able to recover from rare because the points are far away from the beneficial extremes. 
I think this related to the subadditivity criteria of storage effect, i.e., buffered population growth which is related to nonlinear averaging. Subaddivity/ buffered popualtion growth is when species growth rate responses to competition are less affected/sensitive to bad environmental conditions and more affected and sensitive when conditions are good. So it is nonlinear. So points in the beneficial extremes are when the invader can take advantage of the nonlinear averaging that comes with buffered population growth. ATA'd EC relationships have different distribution of points compared to one that is symmetric with the same marginals and correlation so the asymmetry changes whether or not points are in the beneficial extremes.

@@
